---
title: "Illinois Income Tax Impact"
author: "Alex Brunk"
date: "2022-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This will be my example for the Causal Impact model which was developed by Google for doing A/B testing, but which I've used for a variety of program and product evaluation models in my roles at both Ace Hardware Corp and CVS Health.

The purpose of this project is to use the increase in the Illinois income tax in 2011 to illustrate how the causal impact model can be used to evaluate the impact of a one-time change on time series data. In this case, we will attempt to see whether increasing the personal income tax in Illinois had an impact on state GDP.


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(CausalImpact)
library(janitor)
```

First thing we need to do is load the data. You can pull state GDP information by quarter from the Bureau of Economic Analysis. This can be easily downloaded via https://www.bea.gov/data/gdp/gdp-state

```{r, warning=FALSE}
state_gdp_quarter <- read_csv("download.csv",skip = 4)
```

The first row is total US, last three rows are junk and we need to get rid of them. Remove first row and last three rows from dataset

```{r, warning=FALSE}
state_gdp_cleaned <- state_gdp_quarter %>%
  slice(2:(n() - 3))

rownames(state_gdp_cleaned) <- state_gdp_cleaned$GeoName
```

Now lets tranpose to make it easier to work with, get rid of our excess rows with state names and codes, and convert to a data frame with all numeric values.

```{r}
state_gdp_cleaned <- t(state_gdp_cleaned)
state_gdp_cleaned <- state_gdp_cleaned[3:70,]
state_gdp_cleaned <- as.data.frame(state_gdp_cleaned)
state_gdp_cleaned <- state_gdp_cleaned %>%
  mutate_if(is.character,as.numeric)
```

Now, to identify a comparison state, let's create a correlation matrix to identify which states have the strongest correlation to Illinois in GDP growth prior to the tax change. For a pre-period, we will select the period from Jan 2005 through Dec 2010 (the period during which the prior tax rate was in effect per https://www2.illinois.gov/rev/research/taxrates/Pages/individualprioryears.aspx)

```{r}
state_gdp_preperiod <- state_gdp_cleaned[1:24,]
illinois_gdp_preperiod <- state_gdp_preperiod[,14]

state_gdp_correlation <- c()

for(i in 1:ncol(state_gdp_preperiod)) {
  state_gdp_correlation[i] = cor(illinois_gdp_preperiod,state_gdp_preperiod[,i])
}
```

```{r}

correlation_rank <- rbind(state_gdp_preperiod,state_gdp_correlation) %>%
  slice_tail() %>%
  t()

colnames(correlation_rank)[1] = "correlation_perc"
correlation_rank <- as.data.frame(correlation_rank) %>%
  arrange(desc(correlation_perc))

head(correlation_rank,10)
```

Among states with the strongest correlation, South Carolina and Alabama do not have flat income taxes. Indiana on the other hand also has a flat income tax, and made no changes to its tax rates between 2005 and 2014 - see https://taxnews.ey.com/news/2021-1816-indiana-department-of-revenue-announced-county-income-tax-rate-changes-effective-october-1-2021. So we will use Indiana as a comparison state.

A simple chart showing Illinois vs. Indiana for the full 2005 - 2022 time period (note the Covid pandemic beginning in 2020 toward the end)

```{r}

comp_data <- cbind(state_gdp_cleaned[,14],state_gdp_cleaned[,15])

# Visualize the comparison
matplot(comp_data,type="l",ylab='$ GDP',xlab='Quarter',main='Illinois vs. Indiana')

```
<p>Now, using the 2005 to 2010 window for our pre-period, and the 2011 to 2014 window while the increased Illinois income tax was in effect for our post-period, let's see if there appears to have been an impact on state GDP from the higher tax.

To test the impact, we will use the Causal Impact package (documentation: https://google.github.io/CausalImpact/CausalImpact.html) which relies on a Bayesian structural time series model to predict future data in a time series based on past trends.

```{r}
pre.period <- c(1,24)
post.period <- c(25,40)

impact <- CausalImpact(comp_data,pre.period,post.period)

plot(impact)
summary(impact)

```

Counterintuitively, it appears that Illinois's GDP increased relative to Indiana in the years following the tax hike. I don't have a good explanation for a tax hike as a positive causal impact on state GDP. It's possible that the increase in taxes gave investors and businesses confidence that Illinois was solving its economic issues, or that the economy simply improved for other reasons.

<b>As a next step, I may do some additional research to determine if there are other variables that I can account for here. The Causal Impact package allows for the use of a custom bsts model which would allow us to add additional variables to the model other than the quarter.</b>

Either way, this is a simple illustration of how the Causal Impact model can be used to compare time series data.